Plug-and-play dual-tree algorithm runtime analysis

### 2015-08-14 bounds-files/general-runtime-bound.tex
Minor fixes and changes for reviewer #1.

\begin{lemma}
For all nodes $\mathscr{N}_i, \mathscr{N}_j \in R$ (in the context of Algorithm
\ref{alg:cover-tree-dual}) which contain points $p_i$ and $p_j$, respectively,
$d(p_i, p_j) > 2^{s_r^{\max}}$, with $s_r^{\max}$ defined as in line
\ref{alg:line:ct-dual-srmax}.
\end{lemma}

@@

\begin{itemize}
  \item The {\it top scale} $s_q^T$ of the query tree $\mathscr{T}_q$ is such
that as $\lceil \log_2(\eta_q) \rceil - 1 \le s_q^T \le \lceil \log_2(\eta_q)
\rceil$.
  \item The {\it minimum scale} of the query tree $\mathscr{T}_r$ is defined as
$s_q^{\min} = \lceil \log_2(\delta_q) \rceil$.
  \item The top scale $s_r^T$ of the reference tree $\mathscr{T}_r$ is such that
as $\lceil \log_2(\eta_r) \rceil - 1 \le s_r^T \le \lceil \log_2(\eta_r)
\rceil$.
  \item The minimum scale of the reference tree $\mathscr{T}_r$ is defined as
$s_r^{\min} = \lceil \log_2(\delta_r) \rceil$.
\end{itemize}

@@

\begin{equation}
\max\left\{\min\left[O(N \log_2(\delta_q / \delta_r)),
O(N^2)\right], \; 0\right\}.
\end{equation}
\end{lemma}

@@

\begin{equation}
\theta = \max\left\{\min\left[O(N \log_2(\delta_q / \delta_r)), O(N^2)\right],
\; 0\right\}.
\end{equation}
\end{defn}

@@

An intuitive understanding of these bounds is best achieved by first considering
the monochromatic case (this case arises, for instance, in all-nearest-neighbor
search).  The linear dependence on $N$ arises from the fact that all query nodes
must be visited.  The dependence on the reference tree, however, is encapsulated
by the term $c^4 |R^*|$, with $|R^*|$ being the maximum size of the reference
set $R$; this value must be derived for each specific problem.  The bad
performance of poorly-behaved datasets with large $c$ (or, in the worst case, $c
\sim N$) is then captured in both of those terms.  Poorly-behaved datasets may
also have a high cover tree imbalance $I_t(\mathscr{T})$; the linear dependence of
runtime on imbalance is thus sensible for well-behaved datasets.

@@

These results hold for any dual-tree algorithm regardless of the problem. Hence,
%for any pairwise statistical problem, the runtime of the dual-tree algorithm (of
the runtime of any dual-tree algorithm
%that problem)
would be at least $O(N)$ using our bound, which matches the intuition that
answering $O(N)$ queries would take at least $O(N)$ time. For a particular
problem and data, if $c_r$, $|R^*|$, $\chi$, $\psi$ are bounded by constants
independent of $N$ and $\theta$ is no more than linear in $N$ (for large enough
$N$), then the dual-tree algorithm for that problem has a runtime linear in $N$.
Our theoretical result separates out the problem-dependent and the
problem-independent elements of the runtime bound, which allows us to simply
plug in the problem-dependent bounds in order to get runtime bounds for any
dual-tree algorithm without requiring an analysis from scratch.

@@

$\nu$ corresponds to the maximum number of reference recursions between a single
query recursion, and $\kappa$ corresponds to the maximum number of query
recursions between a single reference recursion.  The respective proofs that use
these constants then apply them as a worst-case measure for the whole algorithm:
when using $\kappa$, \citet{ram2009} assume that {\it every} reference recursion
may be followed by $\kappa$ query recursions; similarly, \citet{curtin2014dual}
assume that {\it every} query recursion may be followed by $\nu$ reference
recursions.  Here, we have simply used $I_t(\mathscr{T}_q)$ and $\theta$ as an
exact summation of the total extra reference recursions, which gives us a much
tighter bound than $\nu$ or $\kappa$ on the running time of the whole algorithm.

>>>>

\begin{lemma}
For all distinct nodes $\mathscr{N}_i, \mathscr{N}_j \in R$ (in the context of
Algorithm \ref{alg:cover-tree-dual}) which contain points $p_i$ and $p_j$,
respectively, $d(p_i, p_j) > 2^{s_r^{\max}}$, with $s_r^{\max}$ defined as in
line \ref{alg:line:ct-dual-srmax}.
\end{lemma}

@@

\begin{itemize}
  \item The {\it top scale} $s_q^T$ of the query tree $\mathscr{T}_q$ is such
that as $\lceil \log_2(\eta_q) \rceil - 1 \le s_q^T \le \lceil \log_2(\eta_q)
\rceil$.
  \item The {\it minimum scale} of the query tree $\mathscr{T}_q$ is defined as
$s_q^{\min} = \lceil \log_2(\delta_q) \rceil$.
  \item The top scale $s_r^T$ of the reference tree $\mathscr{T}_r$ is such that
as $\lceil \log_2(\eta_r) \rceil - 1 \le s_r^T \le \lceil \log_2(\eta_r)
\rceil$.
  \item The minimum scale of the reference tree $\mathscr{T}_r$ is defined as
$s_r^{\min} = \lceil \log_2(\delta_r) \rceil$.
\end{itemize}

@@

\begin{equation}
\max\left\{\min\left(O(N \log_2(\delta_q / \delta_r)),
O(N^2)\right), \; 0\right\}.
\end{equation}
\end{lemma}

@@

\begin{equation}
\theta = \max\left\{\min\left(O(N \log_2(\delta_q / \delta_r)), O(N^2)\right),
\; 0\right\}.
\end{equation}
\end{defn}

@@

An intuitive understanding of these bounds is best achieved by first considering
the monochromatic case (this case arises, for instance, in all-nearest-neighbor
search).  The linear dependence on $N$ arises from the fact that all query nodes
must be visited.  The dependence on the reference tree, however, is encapsulated
by the term $c^4 |R^*|$, with $|R^*|$ being the maximum size of the reference
set $R$; this value must be derived for each specific problem.  The poor
performance of trees on datasets with large $c$ (or, in the worst case, $c
\sim N$) is then captured in both of those terms.  These datasets for which
trees perform poorly may also have a high cover tree imbalance
$I_t(\mathscr{T})$; the linear dependence of runtime on imbalance is thus
sensible for datasets where trees perform well.

@@

These results hold for any dual-tree algorithm regardless of the problem. Hence,
%for any pairwise statistical problem, the runtime of the dual-tree algorithm (of
the runtime of any dual-tree algorithm
%that problem)
can be bounded no more tightly than $O(N)$ with our bound, which matches the
intuition that answering $O(N)$ queries will take at least $O(N)$ time. For a
particular problem and data, if $c_r$, $|R^*|$, $\chi$, $\psi$ are bounded by
constants independent of $N$ and $\theta$ is no more than linear in $N$ (for
large enough $N$), then the dual-tree algorithm for that problem has a runtime
linear in $N$.  Our theoretical result separates out the problem-dependent and
the problem-independent elements of the runtime bound, which allows us to simply
plug in the problem-dependent bounds in order to get runtime bounds for any
dual-tree algorithm without requiring an analysis from scratch.

@@

The quantity $\nu$ corresponds to the maximum number of reference recursions
between a single query recursion, and $\kappa$ corresponds to the maximum number
of query recursions between a single reference recursion.  The respective proofs
that use these constants then apply them as a worst-case measure for the whole
algorithm: when using $\kappa$, \citet{ram2009} assume that {\it every}
reference recursion may be followed by $\kappa$ query recursions; similarly,
\citet{curtin2014dual} assume that {\it every} query recursion may be followed
by $\nu$ reference recursions.  Here, we have simply used $I_t(\mathscr{T}_q)$
and $\theta$ as an exact summation of the total extra reference recursions,
which gives us a much tighter bound than $\nu$ or $\kappa$ on the running time
of the whole algorithm.

### 2015-08-14 bounds.tex
Minor fixes and changes for reviewer #1.

A surprising number of machine learning algorithms have computational
bottlenecks that can be expressed as pairwise statistical problems.
By this, we mean computational tasks that can be evaluated directly by iterating over all
pairs of input points.
Nearest neighbor search is one such problem, since for every query point, we can
evaluate its distance to every reference point and keep the closest one.
This naively requires $O(N)$ time
to answer in single query in a reference set of size $N$; answering $O(N)$
queries subsequently requires prohibitive $O(N^2)$ time. Kernel density
estimation is also a pairwise statistical problem, since we compute a sum over all
reference points for each query point.
This again requires $O(N^2)$
time to answer $O(N)$ queries if done directly.
The reference set is typically indexed with
spatial data structures to accelerate this type of computation
\citep{finkel1974quad, langford2006}; these result in $O(\log N)$ runtime per
query under favorable conditions.

>>>>

A surprising number of machine learning algorithms have computational
bottlenecks that can be expressed as pairwise statistical problems.
By this, we mean computational tasks that can be evaluated directly by iterating over all
pairs of input points.
Nearest neighbor search is one such problem, since for every query point, we can
evaluate its distance to every reference point and keep the closest one.
This naively requires $O(N)$ time
to answer a single query in a reference set of size $N$; answering $O(N)$
queries subsequently requires prohibitive $O(N^2)$ time. Kernel density
estimation is also a pairwise statistical problem, since we compute a sum over all
reference points for each query point.
This again requires $O(N^2)$
time to answer $O(N)$ queries if done directly.
The reference set is typically indexed with
spatial data structures to accelerate this type of computation
\citep{finkel1974quad, langford2006}; these result in $O(\log N)$ runtime per
query under favorable conditions.
